A Semantic Proof of Generalised Cut Elimination for Deep Inference

### 2025-05-12 journal-version/old-sections/mav-cut-elimination.tex
[ tex ] work on journal version

\begin{proposition}\label[prop]{prop:embedding-sem}
  For all structures $\vP$, $\sem{\vP} \sqsubseteq \lnot (\ChuEmbed(\ClosedLowerEmbed(P)))$.
\end{proposition}

@@

\begin{theorem}\label[thm]{thm:cut-elim}
  If $\vP\vInferFrom*\vUnit$ in MAV, then there is a normal proof $\vP\vInferFrom*\vUnit$.
\end{theorem}

@@

\begin{theorem}\label[thm]{thm:completeness}
  MAV is complete for the MAV-frame semantics: if, for all MAV-frames
  $F$, $\sem{\vUnit} \sqsubseteq \sem{\vP}$ in
  $(\Chu(\ClosedLowerSet{F}, \ClosedDay{I}), \sqsubseteq)$, then
  $\vP\vInferFrom*\vUnit$.
\end{theorem}

>>>>

\begin{prop}\label[prop]{prop:embedding-sem}
  For all structures $\vP$, $\sem{\vP} \sqsubseteq \lnot (\ChuEmbed(\ClosedLowerEmbed(P)))$.
\end{prop}

@@

\begin{thm}\label[thm]{thm:cut-elim}
  If $\vP\vInferFrom*\vUnit$ in MAV, then there is a normal proof $\vP\vInferFrom*\vUnit$.
\end{thm}

@@

\begin{thm}\label[thm]{thm:completeness}
  MAV is complete for the MAV-frame semantics: if, for all MAV-frames
  $F$, $\sem{\vUnit} \sqsubseteq \sem{\vP}$ in
  $(\Chu(\ClosedLowerSet{F}, \ClosedDay{I}), \sqsubseteq)$, then
  $\vP\vInferFrom*\vUnit$.
\end{thm}

### 2025-05-12 journal-version/old-sections/mav-syntax.tex
[ tex ] work on journal version

\begin{remark}
  In BV, the structural connectives are usually presented as lists, distinguished only by their brackets: $\vP\vTens\vQ$ is written as $\vls(\vP;\vQ)$; $\vP\vParr\vQ$ is written as $\vls[\vP;\vQ]$; and $\vP\vSeq\vQ$ is written as $\vls<\vP;\vQ>$.
  % The equations defining duality are usually included in the decidable equality on structures.
  Inferences, derivations, and proofs are presented vertically, as (\ref{notation:vlin}), (\ref{notation:vlde}), and (\ref{notation:vlpr}), respectively.
  \begin{center}
    $\inlineequation[notation:vlin]{%
        \vlderivation{\vlin{}{}{\vP}{\vlhy{\vQ}}}}$
    \qquad
    $\inlineequation[notation:vlde]{%
        \vlderivation{\vlde{}{}{\vP}{\vlhy{\vQ}}}}$
    \qquad
    $\inlineequation[notation:vlpr]{%
        \vlderivation{\vlpr{}{}{\vP}}}$
  \end{center}

  The relation between the deductive system for BV and rewrite systems is well-known, \eg by Kahramanogullari~\cite{Kahramanogullari08:maude}, who implements proof search for several deep inference systems in Maude~\cite{ClavelDELMMQ02:maude}.
  Inferences rules are usually named with the combination of a letter and an up or down arrow, \eg \cref{rule:Interact} and \cref{rule:CoInteract} are $\iD$ and $\iU$, respectively. The exception are self-dual rules, which are named with a single letter, \eg \cref{rule:Switch} is usually named $\sw$.
\end{remark}

\begin{definition}\label[defn]{defn:normal}
  A derivation is \emph{normal} when it does not use \cref{rule:CoInteract} nor any of the other (\RuleName{Co-}) rules, and its uses of \cref{rule:Interact} are restricted to \cref{rule:AtomInteract}, as defined by the following axiom:
  \begin{displaymath}
    \begin{array}{lcl@{\hspace{0.5cm}}|@{\hspace{0.5cm}}l}
      \makebox[\vInferFromDefnWidthOfLHS][l]{$\vNeg\va\vParr\vPos\va$}
       & \vInferFrom
       & \makebox[\vInferFromDefnWidthOfRHS][l]{$\vUnit$}
       & \RuleLabel{AtomInteract}
    \end{array}
  \end{displaymath}
\end{definition}
Normal derivations avoid the use of rules that introduce new structures into proofs, and so can be termed \emph{analytic} in contrast to the need for the synthetic rules to synthesise new structures.

>>>>

\begin{rem}
  In BV, the structural connectives are usually presented as lists, distinguished only by their brackets: $\vP\vTens\vQ$ is written as $\vls(\vP;\vQ)$; $\vP\vParr\vQ$ is written as $\vls[\vP;\vQ]$; and $\vP\vSeq\vQ$ is written as $\vls<\vP;\vQ>$.
  % The equations defining duality are usually included in the decidable equality on structures.
  Inferences, derivations, and proofs are presented vertically, as (\ref{notation:vlin}), (\ref{notation:vlde}), and (\ref{notation:vlpr}), respectively.
  \begin{center}
    $\inlineequation[notation:vlin]{%
        \vlderivation{\vlin{}{}{\vP}{\vlhy{\vQ}}}}$
    \qquad
    $\inlineequation[notation:vlde]{%
        \vlderivation{\vlde{}{}{\vP}{\vlhy{\vQ}}}}$
    \qquad
    $\inlineequation[notation:vlpr]{%
        \vlderivation{\vlpr{}{}{\vP}}}$
  \end{center}

  The relation between the deductive system for BV and rewrite systems is well-known, \eg by Kahramanogullari~\cite{Kahramanogullari08:maude}, who implements proof search for several deep inference systems in Maude~\cite{ClavelDELMMQ02:maude}.
  Inferences rules are usually named with the combination of a letter and an up or down arrow, \eg \cref{rule:Interact} and \cref{rule:CoInteract} are $\iD$ and $\iU$, respectively. The exception are self-dual rules, which are named with a single letter, \eg \cref{rule:Switch} is usually named $\sw$.
\end{rem}

\begin{defi}\label[defn]{defn:normal}
  A derivation is \emph{normal} when it does not use \cref{rule:CoInteract} nor any of the other (\RuleName{Co-}) rules, and its uses of \cref{rule:Interact} are restricted to \cref{rule:AtomInteract}, as defined by the following axiom:
  \begin{displaymath}
    \begin{array}{lcl@{\hspace{0.5cm}}|@{\hspace{0.5cm}}l}
      \makebox[\vInferFromDefnWidthOfLHS][l]{$\vNeg\va\vParr\vPos\va$}
       & \vInferFrom
       & \makebox[\vInferFromDefnWidthOfRHS][l]{$\vUnit$}
       & \RuleLabel{AtomInteract}
    \end{array}
  \end{displaymath}
\end{defi}
Normal derivations avoid the use of rules that introduce new structures into proofs, and so can be termed \emph{analytic} in contrast to the need for the synthetic rules to synthesise new structures.

### 2025-05-12 journal-version/paper.tex
[ tex ] remove some unneeded package imports
[ tex ] Build the journal version with Nix
[ tex ] work on journal version

\usepackage{hyperref}
\usepackage{mathcommand}
\usepackage{graphicx}
\usepackage{cmll}
\usepackage{mathpartir}
\usepackage[all]{xy}
\usepackage[inline]{enumitem}
\usepackage{amssymb}
\usepackage{fdsymbol}
\usepackage{stmaryrd}
\usepackage{parskip}
\usepackage[inline]{enumitem}
\usepackage{xspace}
\usepackage{multirow}
\usepackage{longtable}
\usepackage[UKenglish,all]{foreign}
\usepackage[goodsyntax]{virginialake}
\vllineartrue%
%\usepackage{url}
%\sloppy

@@

\address{}
\email{me@wen.works}

% FIXME: shorten this
\begin{abstract}
  \emph{Multiplicative-Additive System Virtual} (MAV) is a logic that
  extends Multiplicative-Additive Linear Logic with a self-dual
  non-commutative operator expressing the concept of ``before'' or
  ``sequencing''. MAV is also an extenson of the the logic Basic
  System Virtual (BV) with additives. Formulas in BV have an appealing
  reading as processes with parallel and sequential composition. MAV
  adds internal and external choice operators. BV and MAV are also
  closely related to Concurrent Kleene Algebras.

  Proof systems for MAV and BV are Deep Inference systems, which allow
  inference rules to be applied anywhere inside a structure. As with
  any proof system, a key question is whether proofs in MAV can be
  reduced to a normal form, removing detours and the introduction of
  structures not present in the original goal. In Sequent Calcluli
  systems, this property is referred to as Cut Elimination. Deep
  Inference systems have an analogous Cut rule and other rules that
  are not present in normalised proofs. Cut Elimination for Deep
  Inference systems has the same metatheoretic benefits as for Sequent
  Calculi systems, including consistency and decidability.

  Proofs of Cut Elimination for BV, MAV, and other Deep Inference
  systems present in the literature have relied on intrincate
  syntactic reasoning and complex termination measures.

  We present a concise semantic proof that all MAV proofs can be
  reduced to a normal form avoiding the Cut rule and other ``non
  analytic'' rules. We also develop soundness and completeness proofs
  of MAV (and BV) with respect to a class of models. We have
  mechanised all our proofs in the Agda proof assistant, which
  provides both assurance of their correctness as well as yielding an
  executable normalisation procedure. Our technique extends to include
  exponentials and the additive units.
\end{abstract}

@@

\input{sections/introduction}
%\input{sections/contribution}
\input{sections/mav-syntax}
\input{sections/mav-semantics}
\input{sections/mav-cut-elimination}
\input{sections/mav-mechanisation}
\input{sections/mav-extensions}

>>>>

\usepackage{hyperref}
\usepackage{mathcommand}
\usepackage{cmll}
\usepackage{mathpartir}
\usepackage[inline]{enumitem}
\usepackage{amssymb}
\usepackage{fdsymbol}
\usepackage{stmaryrd}
\usepackage{xspace}
\usepackage{multirow}
\usepackage{longtable}
\usepackage[UKenglish,all]{foreign}
\usepackage[goodsyntax]{virginialake}
\vllineartrue%
%\usepackage{url}
%\sloppy

@@

\address{Well-Typed}
\email{me@wen.works}

% FIXME: shorten this
\begin{abstract}
  We prove cut-elimination properties for some Deep Inference systems
  using semantic techniques.
\end{abstract}

@@

\input{sections/introduction}
\input{sections/mav-syntax}
\input{sections/mav-semantics}
\input{sections/mav-cut-elimination}
\input{sections/mav-mechanisation}
\input{sections/mav-extensions}
